---
title: "Functional enrichment analysis of gene co-expression partners of GATA3"
author: "Gregorio Alanis-Lobato"
output:
  html_document:
    df_print: paged
    fig_caption: yes
---

```{r setup, include=FALSE, message=FALSE}
library(dplyr)
library(readr)
library(cowplot)
library(SingleCellExperiment)
knitr::opts_chunk$set(echo = TRUE)
```

## Intro

In this [R Markdown](http://rmarkdown.rstudio.com) Notebook we perform functional
enrichment analysis of the identified gene co-expression partners of GATA3 
(Yan and Petropoulos integrated dataset).

## GATA3 co-expression partners expressed in the Morula (E4)

First, we load the data:

```{r message=FALSE}
gata3_mor <- read_csv("gata3_partners_expr_in_Morula_YP.csv")
```

Now, we perform a gene ontology (GO) over-representation test of biological 
process annotations, as well as REACTOME pathways, in the set of genes. 

```{r message=FALSE, warning=FALSE}
library(enrichR)

dbs <- c("GO_Biological_Process_2018", "Reactome_2016")

enr <- enrichr(gata3_mor$gene, dbs)
enr$GO_Biological_Process_2018 %>% 
  filter(P.value <= 0.05)

enr$Reactome_2016 %>% 
  filter(P.value <= 0.05)

```

The top-10 most enriched BP terms are:

```{r}
bp <- enr$GO_Biological_Process_2018 %>% 
  arrange(P.value) %>% 
  dplyr::select(Term, P.value, Genes)
bp[1:10, ]
```

The top-10 most enriched REACTOME pathways are:

```{r}
rea <- enr$Reactome_2016 %>% 
  arrange(P.value) %>% 
  dplyr::select(Term, P.value, Genes)
rea[1:10, ]
```
