---
title: "Identification of highly-expressed genes in GATA3- Morula samples"
author: "Gregorio Alanis-Lobato"
output:
  html_document:
    df_print: paged
    fig_caption: yes
---

```{r setup, include=FALSE, message=FALSE}
library(dplyr)
library(readr)
library(cowplot)
library(SingleCellExperiment)
knitr::opts_chunk$set(echo = TRUE)
```

## Intro

In this [R Markdown](http://rmarkdown.rstudio.com) Notebook we identify highly-
expressed genes in Morula samples with low levels of GATA3 expression.

```{r}
load("data/SCE_yan_pet.RData")
```

## Sample selection

We will focus this analysis on the 10% of the Morula samples with the lowest 
levels of GATA3 expression:

```{r gata_lo, echo = FALSE, fig.cap = "**Figure 1.** Expression of GATA3 in Morula samples. The 10% of the cells with the lowest levels of GATA3 are highlighted. GATA3 expression at the blastocyst stage is shown as a reference."}

yp_sce <- yp_sce[, yp_sce$cell_alias == "Morula"]

yp_sce$gata_lo <- logcounts(yp_sce)["GATA3", ] <=
  quantile(logcounts(yp_sce)["GATA3", ], probs = seq(0.1, 1, 0.1))["10%"]

scater::plotExpression(yp_sce, features = "GATA3", x = "cell_alias", 
                       colour_by = "gata_lo", xlab = "", theme_size = 12,
                       show_median = TRUE, add_legend = FALSE)
```

## Marker identification

We now detect genes that are highly expressed in the selected Morula samples. 
These genes show a consistent two-fold higher expression with respect to GATA3
in all samples with low levels of GATA3.

```{r message=FALSE,warning=FALSE}
expr <- logcounts(yp_sce)[, yp_sce$gata_lo]
gata3_avg <- mean(expr["GATA3", ])

fc <- expr/gata3_avg
idx <- rowSums(fc >= 2) == ncol(expr)

genes <- rownames(expr)[idx]
```

The following table lists the identified genes, their average expression in the
GATA3- Morula samples and their average fold-change in expression with respect 
to GATA3 in these cells:

```{r}
tb <- tibble(Gene = genes, `Avg. expression` = rowMeans(expr[idx, ]), 
             `Avg. fold-change` = rowMeans(fc[idx, ])) %>% 
  arrange(desc(`Avg. fold-change`), desc(`Avg. expression`))
tb
```

## Functional enrichment analysis of markers

Finally, we perform a gene ontology (GO) over-representation test of biological 
process annotations, as well as REACTOME pathways, in the set of identified genes. 

```{r message=FALSE, warning=FALSE}
library(enrichR)

dbs <- c("GO_Biological_Process_2018", "Reactome_2016")

enr <- enrichr(genes, dbs)
enr$GO_Biological_Process_2018 %>% 
  filter(P.value <= 0.05)

enr$Reactome_2016 %>% 
  filter(P.value <= 0.05)

```

The top-10 most enriched BP terms are:

```{r}
bp <- enr$GO_Biological_Process_2018 %>% 
  arrange(P.value) %>% 
  dplyr::select(Term, P.value, Genes)
bp[1:10, ]
```

The top-10 most enriched REACTOME pathways are:

```{r}
rea <- enr$Reactome_2016 %>% 
  arrange(P.value) %>% 
  dplyr::select(Term, P.value, Genes)
rea[1:10, ]
```

Since the above tables show that a large number of genes encode for ribosomal
proteins, we remove them from the marker set and re-do the enrichment
analyses:

```{r}
tb <- tb %>% 
  filter(!(grepl("^RP", Gene) | grepl("^MRPS", Gene)))

enr <- enrichr(tb$Gene, dbs)
enr$GO_Biological_Process_2018 %>% 
  filter(P.value <= 0.05)

enr$Reactome_2016 %>% 
  filter(P.value <= 0.05)

tb
```

The top-10 most enriched BP terms are:

```{r}
bp <- enr$GO_Biological_Process_2018 %>% 
  arrange(P.value) %>% 
  dplyr::select(Term, P.value, Genes)
bp[1:10, ]
```

The top-10 most enriched REACTOME pathways are:

```{r}
rea <- enr$Reactome_2016 %>% 
  arrange(P.value) %>% 
  dplyr::select(Term, P.value, Genes)
rea[1:10, ]
```